---
title: "workflow_RNAseq"
author: "Joshua Glowalla"
date: "03292026"
output:
  html_document:
      toc: TRUE #toc = table of contents
      toc_float: TRUE
      toc_depth: 4
      fig_caption: yes
      code_folding: hide
      number_sections: TRUE
vignette: >
    %\VignetteIndexEntry{Text}
    %\usepackage[utf8]{inputenc}
    %\VignetteEngine{knitr::rmarkdown}
fontsize: 13pt
editor_options:
  chunk_output_type: console
---


# NCBI GEO Data

```{r data, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, class.source= "fold-show", results='show', cache=FALSE}

# load base packages and data processing functions for R analysis of biological data
source("Rtools/load_baseline.R")
source("Rtools/load_dataProcessing.R")

# load counts table from GEO
urld <- "https://www.ncbi.nlm.nih.gov/geo/download/?format=file&type=rnaseq_counts"
path <- paste(urld, "acc=GSE81046", "file=GSE81046_raw_counts_GRCh38.p13_NCBI.tsv.gz", sep="&");
tbl <- as.matrix(data.table::fread(path, header=T, colClasses="integer"), rownames=1)

# Load species annotation table from GEO
apath <- paste(urld, "type=rnaseq_counts", "file=Human.GRCh38.p13.annot.tsv.gz", sep="&")
annot <- data.table::fread(apath, header=T, quote="", stringsAsFactors=F, data.table=F)
rownames(annot) <- annot$GeneID

# Load sample annotations table from GEO
sample_table = read.csv("GSE81046_SraRunTable.csv")
table(colnames(tbl) == sample_table$GEO_Accession..exp.)

# removal of gene duplicates 
table(rownames(tbl)==rownames(annot))
table(duplicated(annot$Symbol))
duplicated_indexs = IndexDuplication(annot, "Symbol")
RNAseqCounts = data.frame(tbl[-duplicated_indexs,])
gene_metaData = data.frame(annot[-duplicated_indexs,])
sample_metaData = data.frame(sample_table)


# finalize data format
rownames(RNAseqCounts) = c(gene_metaData$Symbol)
rownames(gene_metaData) = c(gene_metaData$Symbol)

# Remove unnecessary script elements
rm(tbl, annot, sample_table)

```

# Data QC and Filtering

## Code Chunk Part 1

``` {r data_QC1, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, class.source= "fold-show", fig.height=5, fig.width=8, results='show', cache=FALSE}

# == pre-filter low count genes == #
# keep genes with at least 2 counts > 10
keep <- rowSums( RNAseqCounts >= 10 ) >= round(dim(sample_metaData)[1]*0.10)
table(keep)
RNAseqCounts <- RNAseqCounts[keep,]

# == Data Normalization == #
# TPM normalization 
RNAseqCounts_TPM = normTPM(RNAseqCounts, gene_metaData, "Length")

par(mfrow = c(2,2))
# data set too large to render as boxplot so sample indexes are sampled to show goodness of fit
sample_sampling = sample(x = 1:503, size = 100) # first smapling
print("Sample Indexes:")
print(sample_sampling)
boxplot(RNAseqCounts[,sample_sampling], main = "Raw Counts")
boxplot(RNAseqCounts_TPM[,sample_sampling], main = "TPM Normalization")
boxplot(log2(RNAseqCounts[,sample_sampling])+1, col = "lightblue", main = "log2 Raw Counts")
boxplot(log2(RNAseqCounts_TPM[,sample_sampling])+1, col = "lightgreen", main = "log2 Transformed TPM Normalization")
sample_sampling = sample(x = 1:503, size = 100) # second sampling
print("Sample Indexes:")
print(sample_sampling)
boxplot(RNAseqCounts[,sample_sampling], main = "Raw Counts")
boxplot(RNAseqCounts_TPM[,sample_sampling], main = "TPM Normalization")
boxplot(log2(RNAseqCounts[,sample_sampling])+1, col = "lightblue", main = "log2 Raw Counts")
boxplot(log2(RNAseqCounts_TPM[,sample_sampling])+1, col = "lightgreen", main = "log2 Transformed TPM Normalization")
sample_sampling = sample(x = 1:503, size = 100) # third sampling
print("Sample Indexes:")
print(sample_sampling)
boxplot(RNAseqCounts[,sample_sampling], main = "Raw Counts")
boxplot(RNAseqCounts_TPM[,sample_sampling], main = "TPM Normalization")
boxplot(log2(RNAseqCounts[,sample_sampling])+1, col = "lightblue", main = "log2 Raw Counts")
boxplot(log2(RNAseqCounts_TPM[,sample_sampling])+1, col = "lightgreen", main = "log2 Transformed TPM Normalization")

## assessment of variance stabilization of log transformation and TPM normalization
par(mfrow = c(2,2))
varTable = varianceEval(RNAseqCounts, plot_title = "Raw Counts Variance Stability", table_return = TRUE, scatter_return = TRUE)
TPM_varTable = varianceEval(RNAseqCounts_TPM, plot_title = "TPM Normalization Variance Stability", table_return = TRUE, scatter_return = TRUE)
log_varTable = varianceEval(log2(RNAseqCounts)+1, plot_title = "log2 Transformed Raw Counts Variance Stability", table_return = TRUE, scatter_return = TRUE)
logTPM_varTable = varianceEval(log2(RNAseqCounts_TPM)+1, plot_title = "log2 Transformed TPM Normalization Variance Stability", table_return = TRUE, scatter_return = TRUE)
# add gene symbols to data structures for readability
rownames(varTable) = rownames(RNAseqCounts)
rownames(log_varTable) = rownames(RNAseqCounts)
rownames(TPM_varTable) = rownames(RNAseqCounts_TPM)
rownames(logTPM_varTable) = rownames(RNAseqCounts_TPM)
# reorder tables by total variance 
varTable = varTable[order(varTable$variance, decreasing = TRUE),]
log_varTable = log_varTable[order(log_varTable$variance, decreasing = TRUE),]
TPM_varTable = TPM_varTable[order(TPM_varTable$variance, decreasing = TRUE),]
logTPM_varTable = logTPM_varTable[order(logTPM_varTable$variance, decreasing = TRUE),]
# compare top variance genes by normalization and transformation methods
head(varTable, n = 10)
head(TPM_varTable, n = 10)
head(log_varTable, n = 10)
head(logTPM_varTable, n = 10)
# SD vs mean plots to aid in variance stabilization assessment
# library(ggplot2)
# library(gridExtra)
gridExtra::grid.arrange(MSDplot(varTable, "Raw Counts"), MSDplot(log_varTable, "log2 Transformed Raw Counts"), MSDplot(TPM_varTable, "TPM Normalized Counts"), MSDplot(logTPM_varTable, "log2 Transformed TPM Normalized Counts"), ncol = 2, nrow = 2, top = "SD vs Mean Plots")

# transform TPM normalized data
RNAseqCounts_TPM = log2(RNAseqCounts_TPM) + 1
RNAseqCounts_TPM[RNAseqCounts_TPM == Inf | RNAseqCounts_TPM == -Inf] = 0 # remove Inf and -Inf values

# Clear system memory
rm(keep, sample_sampling)
```

## Discussion Part 1

Transformation (log2) provides a significant stabilization of variance across species (genes) for the count data, which is confirmed when comparing the log2 transformed raw counts with the Transcripts per Million (TPM) normalization.  Top variance genes become synonymous between raw counts and TPM normalization after transformation; however, from further analysis of the stabilization of variance across all species improvements to the stabilization is made by the TPM normalization (see 'SD vs Mean Plots').  TPM normalization is a viable method to compare expression profiles because it normalizes by gene length (within sample) and by sequencing depth (between samples) although this method is not considered the most robust way to normalize and 'de-noise' RNAseq data.  To validate and potentially refute this normalization and transformation of the data for downstream analysis, the vst() normalization methods from the DEseq2 package will be evaluated for comparison. rlog() normalization will be overlooked as the sample count is extremely high (503) and this normalization method is better fro low sample counts due to the computational burden of the algorithm. 


``` {r data_QC2, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, class.source= "fold-show", fig.height=5, fig.width=8, results='show', cache=FALSE}

library(DESeq2) # DESeq2 package contains robust normalization methods and differential expression worflow 
rownames(sample_metaData) = colnames(RNAseqCounts) # add rownames to match column names in count matrix
sample_metaData$infection = factor(sample_metaData$infection, levels = c("Non-infected", "Listeria", "Salmonella")) # convert feature varaible under analysis to factors to facilitate differential expression analysis
deSEQ = DESeqDataSetFromMatrix(countData = RNAseqCounts, colData = sample_metaData, design = ~ infection) # DEseq obejct
deSEQ_vst = vst(deSEQ) # VST normalization of raw data
RNAseqCounts_vst = assay(deSEQ_vst) # extract VST nromalized data

# visualize normalization for comparison
par(mfrow= c(2,1))
sample_sampling = sample(x = 1:503, size = 100) # first smapling
print("Sample Indexes:")
print(sample_sampling)
boxplot(RNAseqCounts_TPM[,sample_sampling], col = "lightgreen", main = "log2 Transformed TPM Normalization")
boxplot(RNAseqCounts_vst[,sample_sampling], col = "pink", main = "Varaince Stabilization Normalization")
sample_sampling = sample(x = 1:503, size = 100) # second smapling
print("Sample Indexes:")
print(sample_sampling)
boxplot(RNAseqCounts_TPM[,sample_sampling], col = "lightgreen", main = "log2 Transformed TPM Normalization")
boxplot(RNAseqCounts_vst[,sample_sampling], col = "pink", main = "Varaince Stabilization Normalization")
sample_sampling = sample(x = 1:503, size = 100) # third smapling
print("Sample Indexes:")
print(sample_sampling)
boxplot(RNAseqCounts_TPM[,sample_sampling], col = "lightgreen", main = "log2 Transformed TPM Normalization")
boxplot(RNAseqCounts_vst[,sample_sampling], col = "pink", main = "Varaince Stabilization Normalization")

# Variance Stabilization
par(mfrow = c(1,2))
vst_varTable = varianceEval(RNAseqCounts_vst, plot_title = "VST Normalization", table_return = TRUE, scatter_return = TRUE)
rownames(vst_varTable) = rownames(RNAseqCounts_vst)
vst_varTable = vst_varTable[order(vst_varTable$variance, decreasing = TRUE),]

# Compare SD vs Mean plots between VST and RPM normalization
gridExtra::grid.arrange(MSDplot(logTPM_varTable, "log2 Transformed TPM Normalized Counts"), MSDplot(vst_varTable, "VST Normalization"), nrow = 1, ncol = 2, top = "SD vs Mean Plots: Comparing TPM and VST")

# clear system memory
rm(sample_sampling)

```

## Discussion Part 2

Discuss VSt ver transformed TPM, focus on the trend line and the osilation.  TPM basically has two distinct stabilized groups, arguably three with a transition group between the two.  VST osciliation trend may suggest better stabilization of variance across the range of means between species.  

# Exploratory Analysis by Sample

Performing exploratory analysis by sample is valuable to investigate how well the experimental treatments and/or sample features captures the structure of the expression data.  It is a quality control method as well as validation technique that the assume biological relevance of the experiment is reflected within the data.  The data can undergo cluster analysis/clustering without any dimensional redcution or post-reduction, which helps eliminate noise and expression-level bias prior to clustering.  Ideally, the data feature in question, which represents the experimental design, clusters well and reinforces the pursuit of downstream analysis to find differentiated expressed genes and potential biomarkers. 

## Code Chunk

``` {r EA1_HC, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, class.source= "fold-show", fig.height=5, fig.width=8, results='show', cache=FALSE}
# load EA functions
source("Rtools/load_exploratoryAnalysis.R")

# == Hierarchical Clustering by Sample == #
# bottom-up HC is done by sample to explore how well the biological context groups the samples, distance calculation and linkage is crucial 
# cophenetic distance calculations to optimize HC
cophTable = copheneticEval(count_data = RNAseqCounts_vst, feature_labels = c(t(sample_metaData$infection)))
cophTable
max(cophTable)

# visualize correlation between linkage methods 
print("Linkage Methods:")
print(c("1.ward.D", "2.ward.D2", "3.single", "4.complete", "5.average", "6.mcquitty", "7.median", "8.centroid"))
par(mfrow = c(1,1))
linkageCorrelation(count_data=RNAseqCounts_vst,feature_labels=c(t(sample_metaData$infection)),dist_calculation="Euclidean", plot_title = "Euclidean Distance_VST Normalization Counts")

# visualize and analyze optimized hierarchical clustering 
HC_vst = hclustering_object(count_data = RNAseqCounts_vst, feature_labels = c(t(sample_metaData$infection)), object_type = "dend", distance_type = "Euclidean", linkage_method = "average", dend_plot = TRUE, usage = "Quality Control")
optimizeTree(hc_object = HC_vst, leaf_count = 6, feature_labels = c(sample_metaData$infection))
plot(color_branches(HC_vst, k = 8))

# clear system memory 
# rm(cophTable, HC_vst)
```

## Discussion

Hierarchical clustering was optimized using the cophenetic correlation coefficient and reinforced by pie correlation plot between the different linkage methods for the best distance calculation, euclidean distance. The linkage method "average" was optimal for maximizing the cophenetic correlation coefficient.  To effectively cluster samples by treatment with ~ 98% true postives, 6 clusters was optimal.  Because there are only 3 identified treatment groups, either more relevant (and unintended) biological variation exists within the data or enough noise/expression bias is present to generate significant outliers.  To reveal the root of the observation, dimensional reduction of the data set can be performed to eliminate noise and expression bias between species.  If this approach proves fruitful, it can be safe to conclude that the biological feature in question is robust enough for differential analysis by sample and important species are present driving these differences, which can serve as potential biomarkers for macrophage infection response.  

## Code Chunk

``` {r EA2_PCA, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, class.source= "fold-show", fig.height=5, fig.width=8, results='show', cache=FALSE}
# PCA
#plot()
pca_vst = plotPCA (count_data = RNAseqCounts_vst, feature_labels = c(t(sample_metaData$infection)), plot2D = TRUE, plot3D = FALSE, usage = "sample", return = TRUE)
#ggplot
set.seed(3)
ggplotPCA(count_data = RNAseqCounts_vst, feature_labels = c(t(sample_metaData$infection)), usage = "sample", return = FALSE)

# screeplot
par(mfrow = c(1,1))
screeplot(pca_vst, npcs = 10, type = "barplot", main = "Total Variance Within the First 10 PCs")

## post-PCA hiuerarchical clustering by sample
# revisit optimization
cophTable_pca = copheneticEval(count_data = pca_vst$x[,1:4], feature_labels = c(t(sample_metaData$infection)), usage = "reduction")
cophTable_pca
max(cophTable_pca)
linkageCorrelation(count_data = pca_vst$x[,1:4], feature_labels = c(t(sample_metaData$infection)), dist_calculation="Euclidean", usage = "reduction", plot_title = "Euclidean Distance_PCA reduction (PC1-4)")

# perfrom hierarchical clustering by sample
pcaHC_vst = hclustering_object(count_data = pca_vst$x[,1:4], feature_labels = c(t(sample_metaData$infection)), object_type = "dend", distance_type = "Euclidean", linkage_method = "average", dend_plot = TRUE, usage = "reduction")
table(sample_metaData$infection) # number of samples by infection type
optimizeTree(hc_object = pcaHC_vst, leaf_count = 6, feature_labels = c(sample_metaData$infection)) # reports which samples by feature are clustered within each cluster
plot(color_branches(pcaHC_vst, k = 3)) # plot optimal clustering number found from analysis

# clear system memory
set.seed(NULL)
rm()
```

## Discussion

PCA was performed to reduce the dimension of the count data set.  Covariance captured within PC1 aligns perfectly with the experimental treatment.  PC2 captures variance in a dimension that does not exhibit any alignment with infection type while PC3 does well at separating infection from non-infection samples, suggesting that the variance in this dimension has some sort of biological relevance in line with the experiment.  A screeplot identifies that the overwhelming majority of the covariance within the data is captured within the first 4 PCs.  After truncating the data along with the dimensional reduction of PCA (use of PCs 1-4), outliers from previous hierarchical clustering by samples were avoided and were able to cluster the groups to experimental infection treatments with 100% accuracy and reflect the PCA results interpretation. Other dimensional reduction techniques are viable for reducing the noise and improving the expression bias.  TO follow up, t-SNE and UMAp methods will be optimized and then evaluated.  

## Code Chunk

``` {r EA3_tSNE_UMAP, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, class.source= "fold-show", fig.height=5, fig.width=8, results='show', cache=FALSE}

# t-SNE optimization along with parameter tuning for hierarchical and kmeans clustering 
tsneOptimize(data_object = t(RNAseqCounts_vst), feature_varaible = c(sample_metaData$infection))
tsneOptimize(data_object = t(RNAseqCounts_vst), feature_varaible = c(sample_metaData$infection), usage = "check", neighbors = 28)
tSNE_samples = Rtsne(t(RNAseqCounts_vst), dims = 2, perplexity = 28, pca = TRUE, theta = 0.25, check_duplicates = FALSE)
# hclustering 
cophTable_tsne = copheneticEval(data_object = tSNE_samples$Y, feature_labels = c(t(sample_metaData$infection)), usage = "reduction")
cophTable_tsne = copheneticEval2(data_object = tSNE_samples$Y, transpose_data = "features")
cophTable_tsne
# kmeans
kclusterOptimize(tSNE_samples$Y, cluster_range = 10, iteration_max = 10, kmean_algorithm = "Hartigan-Wong")
kclusterWithinSS(tSNE_samples$Y, cluster_count = 3, normalize_comparison = TRUE)

# UMAP optimization along with parameter tuning for hierarchical and kmeans clustering 
umapOptimize(data_object = t(RNAseqCounts_vst), feature_varaible = c(sample_metaData$infection))
umapOptimize(data_object = t(RNAseqCounts_vst), feature_varaible = c(sample_metaData$infection), usage = "check", neighbors = 20)
umap_samples = umap(t(RNAseqCounts_vst), method = "naive", n_neighbors = 20, n_components = 2, min_dist = 0.05, metric = "euclidean")
# hclustering 
cophTable_umap = copheneticEval(count_data = umap_samples$layout, feature_labels = c(t(sample_metaData$infection)), usage = "reduction")
cophTable_umap
# kmeans
kclusterOptimize(umap_samples$layout, cluster_range = 10, iteration_max = 10, kmean_algorithm = "Hartigan-Wong")
kclusterWithinSS(umap_samples$layout, cluster_count = 3, normalize_comparison = TRUE)

```

## Discussion

Successful optmimization of dimensional reduction methods t-SNE and UMAP to well represent the treatment groups.  CLuster analysis must be performed to cooraborate the results for all three methods.  

## Code Chunk

``` {r EA3_FinalAnaly sis, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, class.source= "fold-show", fig.height=5, fig.width=8, results='show', cache=FALSE}

# PCA Dimensional Reduction for Clustering Algorithms (Hierarchical and Kmeans)
clusterAnalysis(data_object = RNAseqCounts_vst, feature_labels = c(sample_metaData$infection), cluster_count = 6, dimension_reduction = "pca", cluster_algorithm = "hclust", algorithm_method = "average", dist_method = "euclidean", transpose_data = TRUE, dim_output = 2, local_global = 28, dist_min = 0.05)
clusterAnalysis(data_object = RNAseqCounts_vst, feature_labels = c(sample_metaData$infection), cluster_count = 3, dimension_reduction = "pca", cluster_algorithm = "kmeans", algorithm_method = "Hartigan-Wong", dist_method = "euclidean", transpose_data = TRUE, dim_output = 2, local_global = 28, dist_min = 0.05)

# tSNE Dimensional Reduction for Clustering Algorithms (Hierarchical and Kmeans)
clusterAnalysis(data_object = RNAseqCounts_vst, feature_labels = c(sample_metaData$infection), cluster_count = 3, dimension_reduction = "tsne", cluster_algorithm = "hclust", algorithm_method = "average", dist_method = "euclidean", transpose_data = TRUE, dim_output = 2, local_global = 28, dist_min = 0.05)
clusterAnalysis(data_object = RNAseqCounts_vst, feature_labels = c(sample_metaData$infection), cluster_count = 3, dimension_reduction = "tsne", cluster_algorithm = "kmeans", algorithm_method = "Hartigan-Wong", dist_method = "euclidean", transpose_data = TRUE, dim_output = 2, local_global = 28, dist_min = 0.05)

# UMAP Dimensional Reduction for Clustering Algorithms (Hierarchical and Kmeans)
clusterAnalysis(data_object = RNAseqCounts_vst, feature_labels = c(sample_metaData$infection), cluster_count = 3, dimension_reduction = "umap", cluster_algorithm = "hclust", algorithm_method = "average", dist_method = "euclidean", transpose_data = TRUE, dim_output = 2, local_global = 20, dist_min = 0.05)
clusterAnalysis(data_object = RNAseqCounts_vst, feature_labels = c(sample_metaData$infection), cluster_count = 3, dimension_reduction = "umap", cluster_algorithm = "kmeans", algorithm_method = "Hartigan-Wong", dist_method = "euclidean", transpose_data = TRUE, dim_output = 2, local_global = 20, dist_min = 0.05)

```

## Discussion

Along with dimensionality reduction, clustering successfully captures the biological feature (macrophage infection response).  Reduction through PCA, hierarchical clustering was unable to match cluster number with infection type as several outliers distort the clustering.  At 6 perceived clusters, the bulk of the separation was finally captured via clustering.  Reduction through UMAP provides tighter clusters while tSNE generates clusters with a significantly wider spread. However, both algorithms still provide adequate inter-cluster spread. Each of the techniques was successful in matching the cluster number with infection type.  Based on these results infection type is captured well by the gene expression differences between samples.  Therefore, exploratory analysis by species (genes) will yield the genomic contribution to the phenotypic differences of macrophages by infection type.  

# Exploratory Analysis by Genes

## Code Chunk

``` {r explore_gene, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, class.source= "fold-show", fig.height=5, fig.width=8, results='show', cache=FALSE}
varTable_topVar = vst_varTable[vst_varTable$variance>3,]

RNAseqCounts_topVar = RNAseqCounts_vst[rownames(varTable_topVar),]

# top 1000 genes by variance > 1
pca_genes = plotPCA(count_data = RNAseqCounts_topVar, feature_labels = rownames(RNAseqCounts_topVar), usage = "species", plot2D = TRUE, plot3D = FALSE, return = TRUE)
sAnalyzePCA(pcaData_1 = pca_vst , pcaData2 = pca_genes, species_labels = rownames(RNAseqCounts_topVar), main_title = "PCA for Top Loading Genes", input_type = "pca", visualize = TRUE, visual_count = 5, return = "genes")

# screeplot
par(mfrow = c(1,1))
screeplot(pca_genes, npcs = 10, type = "barplot", main = "Total Variance Within the First 10 PCs")

cophTable_genePCA = copheneticEval(data_object = pca_genes$x[,1:3], feature_labels = rownames(RNAseqCounts_vst), usage = "reduction")
cophTable_genePCA

kclusterOptimize(pca_genes$x[,1:3], cluster_range = 20, iteration_max = 10, kmean_algorithm = "Hartigan-Wong")
kclusterWithinSS(pca_genes$x[,1:3], cluster_count = 6, normalize_comparison = TRUE)


# t-SNE optimization along with parameter tuning for hierarchical and kmeans clustering 
tsneOptimize(data_object = RNAseqCounts_topVar, feature_varaible = rownames(RNAseqCounts_vst), duplicates = FALSE)
tsneOptimize(data_object = RNAseqCounts_topVar, feature_varaible = rownames(RNAseqCounts_vst), usage = "check", neighbors = 12, duplicates = FALSE)
tSNE_genes = Rtsne(RNAseqCounts_topVar, dims = 2, perplexity = 12, pca = TRUE, theta = 0.25, check_duplicates = FALSE)
# hclustering 
cophTable_tsne = copheneticEval(count_data = tSNE_genes$Y, feature_labels = rownames(RNAseqCounts_vst), usage = "reduction")
cophTable_tsne = copheneticEval2(data_object = tSNE_genes$Y, transpose_data = "species")
cophTable_tsne
# kmeans
kclusterOptimize(tSNE_samples$Y, cluster_range = 10, iteration_max = 10, kmean_algorithm = "Hartigan-Wong")
kclusterWithinSS(tSNE_samples$Y, cluster_count = 5, normalize_comparison = TRUE)

# UMAP optimization along with parameter tuning for hierarchical and kmeans clustering 
umapOptimize(data_object = RNAseqCounts_topVar, feature_varaible = rownames(RNAseqCounts_vst))
umapOptimize(data_object = RNAseqCounts_topVar, feature_varaible = rownames(RNAseqCounts_vst), usage = "check", neighbors = 20)
umap_genes = umap(RNAseqCounts_topVar, method = "naive", n_neighbors = 20, n_components = 2, min_dist = 0.05, metric = "euclidean")
# hclustering 
cophTable_umap = copheneticEval(count_data = umap_genes$layout, feature_labels = rownames(y), usage = "reduction")
cophTable_umap
# kmeans
kclusterOptimize(umap_samples$layout, cluster_range = 10, iteration_max = 10, kmean_algorithm = "Hartigan-Wong")
kclusterWithinSS(umap_samples$layout, cluster_count = 6, normalize_comparison = TRUE)


# PCA Dimensional Reduction for Clustering Algorithms (Hierarchical and Kmeans)
clusterAnalysis2(data_object = pca_genes, feature_labels = rownames(RNAseqCounts_topVar), cluster_count = 6, dimension_reduction = "pca", cluster_algorithm = "hclust", algorithm_method = "average", dist_method = "euclidean", transpose_data = FALSE, use_dim = 10, return = FALSE)
clusterAnalysis2(data_object = pca_genes, feature_labels = rownames(RNAseqCounts_topVar), cluster_count = 6, dimension_reduction = "pca", cluster_algorithm = "kmeans", algorithm_method = "Hartigan-Wong", dist_method = "euclidean", transpose_data = FALSE, use_dim = 10, return = FALSE)





clusterAnalysis(data_object = RNAseqCounts_topVar, feature_labels = rownames(RNAseqCounts_topVar), cluster_count = 6, dimension_reduction = "pca", cluster_algorithm = "hclust", algorithm_method = "average", dist_method = "Pearson", transpose_data = FALSE, dim_output = 2, local_global = 6, dist_min = 0.05)
clusterAnalysis(data_object = RNAseqCounts_topVar, feature_labels = rownames(RNAseqCounts_topVar), cluster_count = 6, dimension_reduction = "pca", cluster_algorithm = "kmeans", algorithm_method = "Hartigan-Wong", dist_method = "euclidean", transpose_data = FALSE, dim_output = 2, local_global = 6, dist_min = 0.05)

# tSNE Dimensional Reduction for Clustering Algorithms (Hierarchical and Kmeans)
clusterAnalysis(data_object = RNAseqCounts_topVar, feature_labels = rownames(RNAseqCounts_topVar), cluster_count = 3, dimension_reduction = "tsne", cluster_algorithm = "hclust", algorithm_method = "average", dist_method = "pearson", transpose_data = FALSE, dim_output = 2, local_global = 12, dist_min = 0.05)
clusterAnalysis(data_object = RNAseqCounts_topVar, feature_labels = rownames(RNAseqCounts_topVar), cluster_count = 5, dimension_reduction = "tsne", cluster_algorithm = "kmeans", algorithm_method = "Hartigan-Wong", dist_method = "euclidean", transpose_data = FALSE, dim_output = 2, local_global = 12, dist_min = 0.05)

# UMAP Dimensional Reduction for Clustering Algorithms (Hierarchical and Kmeans)
clusterAnalysis(data_object = RNAseqCounts_topVar, feature_labels = rownames(RNAseqCounts_topVar), cluster_count = 3, dimension_reduction = "umap", cluster_algorithm = "hclust", algorithm_method = "average", dist_method = "euclidean", transpose_data = FALSE, dim_output = 2, local_global = 6, dist_min = 0.05, return = "none")
clusterAnalysis(data_object = RNAseqCounts_topVar, feature_labels = rownames(RNAseqCounts_topVar), cluster_count = 3, dimension_reduction = "umap", cluster_algorithm = "kmeans", algorithm_method = "Hartigan-Wong", dist_method = "euclidean", transpose_data = FALSE, dim_output = 2, local_global = 6, dist_min = 0.05)

```

## Discussion


## Code Chunk

``` {r Q2, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, class.source= "fold-show", fig.height=5, fig.width=8, results='show', cache=FALSE}
# PCA



```

## Discussion

## Discussion

## Code Chunk

``` {r EA3_, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, class.source= "fold-show", fig.height=5, fig.width=8, results='show', cache=FALSE}


```

## Discussion

# Section Title

## Code Chunk

``` {r Q2, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, class.source= "fold-show", fig.height=5, fig.width=8, results='show', cache=FALSE}

w = hclustObject(data_object = tSNE_genes$Y, dist_method = "euclidean", linkage_method = "average", transpose_data = "species", verbose = TRUE)
x = hclustObject(data_object = tSNE_genes$Y, dist_method = "pearson", linkage_method = "ward.D2", transpose_data = "species", verbose = TRUE)
xd = as.dist(1-stats::cor(t(tSNE_genes$Y), method = "pearson"))
y
z


```

## Discussion



# Question 3

## Code Chunk

``` {r Q3, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, class.source= "fold-show", fig.height=5, fig.width=8, results='show', cache=FALSE}
```

## Discussion



# Question 4

## Code Chunk

``` {r Q4, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, class.source= "fold-show", fig.height=5, fig.width=8, results='show', cache=FALSE}
```

## Discussion