---
title: "workflow_RNAseq"
author: "Joshua Glowalla"
date: "03292026"
output:
  html_document:
      toc: TRUE #toc = table of contents
      toc_float: TRUE
      toc_depth: 4
      fig_caption: yes
      code_folding: hide
      number_sections: TRUE
vignette: >
    %\VignetteIndexEntry{Text}
    %\usepackage[utf8]{inputenc}
    %\VignetteEngine{knitr::rmarkdown}
fontsize: 13pt
editor_options:
  chunk_output_type: console
---


# NCBI GEO Data

## GEO Data Code Chunl

```{r data, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, class.source= "fold-show", results='show', cache=FALSE}

# load base packages and data processing functions for R analysis of biological data
source("Rtools/load_baseline.R")
source("Rtools/load_dataProcessing.R")

# load counts table from GEO
urld <- "https://www.ncbi.nlm.nih.gov/geo/download/?format=file&type=rnaseq_counts"
path <- paste(urld, "acc=GSE81046", "file=GSE81046_raw_counts_GRCh38.p13_NCBI.tsv.gz", sep="&");
tbl <- as.matrix(data.table::fread(path, header=T, colClasses="integer"), rownames=1)

# Load species annotation table from GEO
apath <- paste(urld, "type=rnaseq_counts", "file=Human.GRCh38.p13.annot.tsv.gz", sep="&")
annot <- data.table::fread(apath, header=T, quote="", stringsAsFactors=F, data.table=F)
rownames(annot) <- annot$GeneID

# Load sample annotations table from GEO
setwd("Desktop/Github Projects/Project_Portfolio_JMG")
sample_table = read.csv("GSE81046_SraRunTable.csv")
table(colnames(tbl) == sample_table$GEO_Accession..exp.)

# removal of gene duplicates 
table(rownames(tbl)==rownames(annot))
table(duplicated(annot$Symbol))
duplicated_indexs = IndexDuplication(annot, "Symbol")
RNAseqCounts = data.frame(tbl[-duplicated_indexs,])
gene_metaData = data.frame(annot[-duplicated_indexs,])
sample_metaData = data.frame(sample_table)
rownames(sample_metaData) = sample_metaData$Sample.Name

# Ancestry data missing from sample annotations
ancestry_table = as.data.frame(read.csv("GSE81046_ancestry.csv"))
rownames(ancestry_table) = ancestry_table[,"sample_ID"]
ancestry_table = ancestry_table[rownames(sample_metaData),]


# finalize data structures
rownames(RNAseqCounts) = c(gene_metaData$Symbol)
rownames(gene_metaData) = c(gene_metaData$Symbol)
sample_metaData$ancestry_ID = ancestry_table$ancestry_ID
sample_metaData$ancestry = codeLabel(feature_labels = c(sample_metaData$ancestry_ID), label_pattern = c("AF","EU"), label_match = c("African","European"))
sample_metaData$infection_ancestry = codeFeature(data_object = sample_metaData, feature_cols = c("infection","ancestry"))

table(sample_metaData$ancestry)
table(sample_metaData$infection)
table(sample_metaData$infection_ancestry)

# Remove unnecessary script elements
rm(tbl, annot, sample_table, ancestary_table)

```

## GEO Data Discussion

Research published out of the University of Chicago and uploaded by Luis Barreiro in 2016 produced a public bulk RNAseq data set intent on investigating the impact of genetic ancestry on innate immune response to pathogenic insult, specifically the two common food poisoning bacterium Listeria monocytogenes (Listeria) and Salmonella enterica (Salmonella).  Primary macrophages from African and European descent we tested against these pathogens and their transcriptomic profiles after 2 hours of either no infection, listeria or salmonella infections were presented for a total of 503 samples split amoung the ancestry and infection groups. 


** add a review of the results with citations as a benchmark for the results obtained in this


# Data QC and Filtering

## QC1 Code Chunk

``` {r QC1, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, class.source= "fold-show", fig.height=5, fig.width=8, results='show', cache=FALSE}

# == pre-filter low count genes == #
# keep genes with at least CPM = 1 in (# of samples in smallest experimental group) 
RNAseqCounts = filterCPM(data_object = RNAseqCounts, feature_labels = c(sample_metaData$infection))
gene_metaData = gene_metaData[rownames(RNAseqCounts),]
# keep genes with at least 2 counts > 10
#keep <- rowSums( RNAseqCounts >= 10 ) >= round(dim(sample_metaData)[1]*0.10)
#table(keep)
#RNAseqCounts <- RNAseqCounts[keep,]

# == Data Normalization == #
# TPM normalization 
RNAseqCounts_TPM = normTPM(count_data = RNAseqCounts, gene_data = gene_metaData, col_name = "Length")

# define default graphical parameters
old_par = par(no.readonly = TRUE)

par(mfrow = c(2,2))
# data set too large to render as box plot so sample indexes are sampled to show goodness of fit
sample_sampling = sample(x = 1:503, size = 100) # first sampling
print("Sample Indexes:")
print(sample_sampling)
boxplot(RNAseqCounts[,sample_sampling], main = "Raw Counts")
boxplot(RNAseqCounts_TPM[,sample_sampling], main = "TPM Normalization")
boxplot(log2(RNAseqCounts[,sample_sampling])+1, col = "lightblue", main = "log2 Raw Counts")
boxplot(log2(RNAseqCounts_TPM[,sample_sampling])+1, col = "lightgreen", main = "log2 Transformed TPM Normalization")
sample_sampling = sample(x = 1:503, size = 100) # second sampling
print("Sample Indexes:")
print(sample_sampling)
boxplot(RNAseqCounts[,sample_sampling], main = "Raw Counts")
boxplot(RNAseqCounts_TPM[,sample_sampling], main = "TPM Normalization")
boxplot(log2(RNAseqCounts[,sample_sampling])+1, col = "lightblue", main = "log2 Raw Counts")
boxplot(log2(RNAseqCounts_TPM[,sample_sampling])+1, col = "lightgreen", main = "log2 Transformed TPM Normalization")
sample_sampling = sample(x = 1:503, size = 100) # third sampling
print("Sample Indexes:")
print(sample_sampling)
boxplot(RNAseqCounts[,sample_sampling], main = "Raw Counts")
boxplot(RNAseqCounts_TPM[,sample_sampling], main = "TPM Normalization")
boxplot(log2(RNAseqCounts[,sample_sampling])+1, col = "lightblue", main = "log2 Raw Counts")
boxplot(log2(RNAseqCounts_TPM[,sample_sampling])+1, col = "lightgreen", main = "log2 Transformed TPM Normalization")

# assessment of data quality by variance stability
par(old_par)
par(mfrow = c(1,2))
# raw data
summary_RAW = analysisVariance(data_object = RNAseqCounts, plot_title = "Raw Data Variance Stability", data_return = "summary")
# TPM normalization
summary_TPM = analysisVariance(data_object = RNAseqCounts_TPM, plot_title = "TPM Normalization Variance Stability", data_return = "summary")
# log2 transformed raw data 
summary_logRAW = analysisVariance(data_object = log2(RNAseqCounts)+1, plot_title = "Transformed Raw Data Variance Stability", data_return = "summary")
# log2 transformed TPM
summary_logTPM = analysisVariance(data_object = log2(RNAseqCounts_TPM)+1, plot_title = "Transformed TPM Variance Stability", data_return = "summary")

## reorder tables by total variance 
summary_RAW = summary_RAW[order(summary_RAW$variance, decreasing = TRUE),]
summary_logRAW = summary_logRAW[order(summary_logRAW$variance, decreasing = TRUE),]
summary_TPM = summary_TPM[order(summary_TPM$variance, decreasing = TRUE),]
summary_logTPM = summary_logTPM[order(summary_logTPM$variance, decreasing = TRUE),]

# compare top 10 genes by variance for each normalization and transformation method
head(summary_RAW, n = 10)
head(summary_logRAW, n = 10)
head(summary_TPM, n = 10)
head(summary_logTPM, n = 10)

# finalize transformation of TPM normalized data
RNAseqCounts_TPM = log2(RNAseqCounts_TPM) + 1
RNAseqCounts_TPM[RNAseqCounts_TPM == Inf | RNAseqCounts_TPM == -Inf] = 0 # remove Inf and -Inf values

# Clear system memory
par(old_par)
rm(sample_sampling)
```

## QC1 Discussion

First stage of quality control is to filter out low count species.  Although a less stringent filter is commonly applied, total counts across samples > 10, best practice is to use sequencing depth to normalize the data and remove species  where total number of samples with counts per million equal to or greater than one is less than the number of samples within the smallest treatment size.  The incorporation of sample depth allows for count sizes between samples to be better comparable and not bias samples with more sequencing depth, which is the case with raw read counts.  Where a fixed total read counts across samples biases total sample count for the data, the CPM normalization and smallest treatment size count generates a more robust low count species filtering process applicable to any data set to effectively remove these troublesome species for downstream analysis.

Next, transformations (log2) and normalization of the data is evaluated for improving sample-by-sample comparison.  Although CPM is a normalization method, it does not offer a within-sample normalization step and would not solve a read count bias for the species-by-species comparisons in the analytic process.  Instead transcripts per million (TPM) is applied as the primary normalization method, which counts for biases in sequencing depth and transcript length.  Key analysis of this stage requires evaluating the distributions of read counts between samples and species.  Box-whisker plots between samples provide a strong visualization tool to view how well a method normalizes the overall samples to each other and the Mean vs SD plots (MSD plots) provide tool to see how well read counts distributions between species is stabilized.  The methods applied here, however, are not variance stabilization methods and although the results suggest that TPM normalization and log2 transformation jointly provides decent stabilization of variance and distribution of reads by sample other method exist with this aim.  

To validate or refute use of TPM normalization and log2 transformation of the data for downstream analysis, the vst() normalization method from the DEseq2 package will be evaluated for comparison. rlog() normalization will be overlooked as the sample count is extremely high (503) and rlog() is better suited for low sample counts due to the computational burden of the algorithm. 


## QC2 Data Chunk

``` {r QC2, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, class.source= "fold-show", fig.height=5, fig.width=8, results='show', cache=FALSE}

library(DESeq2) # DESeq2 package contains robust normalization methods and differential expression worflow 
rownames(sample_metaData) = colnames(RNAseqCounts) # add rownames to match column names in count matrix
# convert feature variable under analysis to factors to facilitate differential expression analysis
sample_metaData$infection = factor(sample_metaData$infection, levels = c("Non-infected", "Listeria", "Salmonella")) 
sample_metaData$ancestry = factor(sample_metaData$ancestry, levels = c("African","European"))
sample_metaData$infection_ancestry = factor(sample_metaData$infection_ancestry, levels = c("Non-infected_African","Non-infected_European","Listeria_African","Listeria_European","Salmonella_African","Salmonella_European"))
deSEQ = DESeqDataSetFromMatrix(countData = RNAseqCounts, colData = sample_metaData, design = ~ infection_ancestry) # DEseq obejct
deSEQ_vst = vst(deSEQ) # VST normalization of raw data
RNAseqCounts_vst = assay(deSEQ_vst) # extract VST nromalized data

# visualize normalization for comparison
par(mfrow= c(2,1))
sample_sampling = sample(x = 1:503, size = 100) # first smapling
print("Sample Indexes:")
print(sample_sampling)
boxplot(RNAseqCounts_TPM[,sample_sampling], col = "lightgreen", main = "log2 Transformed TPM Normalization")
boxplot(RNAseqCounts_vst[,sample_sampling], col = "pink", main = "Varaince Stabilization Normalization")
sample_sampling = sample(x = 1:503, size = 100) # second smapling
print("Sample Indexes:")
print(sample_sampling)
boxplot(RNAseqCounts_TPM[,sample_sampling], col = "lightgreen", main = "log2 Transformed TPM Normalization")
boxplot(RNAseqCounts_vst[,sample_sampling], col = "pink", main = "Varaince Stabilization Normalization")
sample_sampling = sample(x = 1:503, size = 100) # third smapling
print("Sample Indexes:")
print(sample_sampling)
boxplot(RNAseqCounts_TPM[,sample_sampling], col = "lightgreen", main = "log2 Transformed TPM Normalization")
boxplot(RNAseqCounts_vst[,sample_sampling], col = "pink", main = "Varaince Stabilization Normalization")

# Variance Stabilization
par(mfrow = c(1,2))
summary_vst = analysisVariance(data_object = RNAseqCounts_vst, plot_title = "VST Normalization", data_return = "summary")
summary_vst = summary_vst[order(summary_vst$variance, decreasing = TRUE),]
head(summary_vst, n = 10)

# clear system memory
par(old_par)
rm(sample_sampling)
```

## QC2 Discussion

DESeq2 package provides the vst() function ("variance stabilization transformation").  As expected this normalization method improves read count distribution across all samples and variance across the range of mean read counts by species.  The MSD plot for the TPM normalization reveals two, potentially three, distinct groups defined by the mean and each stabilized at a different variance.  The VST normalization displays a minimal oscillating variance average across the range of mean read counts. Although neither is the ideal trend line, the VST normalization range of average variance is less than that of the TPM.  Also, the lack of a clear pattern across the range of means suggests a more successful normalization rather than still containing an artifact of the inherent bias in species with higher or lower expression.  Variance is not truly stabilized but the species with higher or lower variance can not be clearly defined, especially by their mean expression.  It is with this rationale that the VST normalization will move forward in the analytic pipeline. 

# Unsupervised Learning (Exploratory Analysis): Dimensionality Reduction

Dimensionality reduction (DR) methods are powerful tools to reduce data with high dimensions to a smaller number of dimensions, possibly just two, while preserving the global and local relationships between data points in the original data.  Within this reduced dimensional space much of the noise can be removed, global anomalies and local outliers can be identified and features of interest are capable of being analyzed visually, which is at the very least extremely difficult if not impossible within the space of the original data.  The former is crucial for finalizing data QC while the latter is a powerful tool to both support the biological question being asked and facilitate the generation of targeted, testable hypotheses.  To successfully reach these goals, it is often required to perform an analysis of the DR and/or undergo hyper parameter tuning to optimize the DR.  The most common DR methods are PCA (Principal Component Analysis), t-SNE (t-Distributed Stochastic Neighbor Embedding), UMAP (Uniform Manifold Approximation and Projection), and MDS (Multidimensional Scaling).  PCA is a linear approach that identifies dimensions in order of greatest variance (Principal Components) - i.e. each subsequent dimension (axes) houses less variance between the data points than the last. Both t-SNE and UMAP are nonlinear DR methods that allow the user to define the number of dimensions the data will be reduced to and excel at capturing local and global structures within the data.  t-SNE is known better for capturing local structure and therefore excellent for scRNAseq data analysis while UMAP can be preferred over t-SNE since it is computationally less demanding (faster) and considered to better balanced for capture of global and local structure. Within this context, PCA would be considered a strong DR technique to evaluate global structure.  MDS focuses on preserving pairwise distances between data points - i.e. it maps the data into reduced dimensions in which the more similar data is the closer they are to each other within this space.  This is uniquely different from t-SNE and UMAP, which the distances of data within these reduced dimensional space is not as strongly connected to similarity and "tunable".  This difference makes MDS a strong DR method for pre-processing prior to K-means or DBSCAN clustering algorithms as well as outlier detection. Following DR, the data in the reduced dimensional space can be applied to unsupervised learning approaches - i.e. clustering - in order to finish the data quality control process and investigate the biology of interest as stated above.  

Before several unsupervised learning methods including various DR tools are explored, an example of applying unsupervised learning to normalized and transformed data without DR will help highlight the advantages of DR thereafter.  To do this, hierarchical clustering will be applied to the data set and observe how well the feature structure is captured by the approach as a baseline for comparison.  

## An Example of Unsupervised Learning without Dimensionality Reduction of High Dimensional Data

###  EA1 Code Chunk

``` {r EA1, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, class.source= "fold-show", fig.height=5, fig.width=8, results='show', cache=FALSE}

# load functions (tools) for exploratory analysis
source("Rtools/load_exploratoryAnalysis.R")

# Hierarchical clustering of the normalized and transformed data set
# Cophenetic correlation coefficient to evaluate global accuracy of hclust object structure for combinations of distance calculations and linkage methods
# Limits of cophenetic correlation coefficients can be mitigated by correlation plots between them

cophTable = analysisCophenetic(data_object = RNAseqCounts_vst, feature_labels = c(sample_metaData$infection_ancestry)) # combinatorial evaluation of cophenetic correlation coefficients
cophTable

par(old_par)
analysisHclustCor(data_object = RNAseqCounts_vst, feature_labels = c(sample_metaData$infection_ancestry),dist_calculation = c("euclidean","pearson")) # correlations between these the best distance calculation methods
dend_vst = objectHCLUST(data_object = RNAseqCounts_vst, feature_labels = c(sample_metaData$infection_ancestry), return_object = "dend", dend_plot = TRUE) # view dendrogram of optimal distance calculation and linkage method and create dendrogram object
optimizeTree(hc_object = dend_vst, leaf_count = 10, feature_labels = c(sample_metaData$infection_ancestry)) # evaluate cutting of the structure into a set nu8mber of groups and see how well it clusters the data by feature
plot(color_branches(dend_vst, k = 6)) # view dendrogram with the optimal number of leaves to capture the biology

# clear system memory 
# rm(cophTable, dend_vst)

```

### EA1 Discussion

Because hierarchical clustering requires data to be converted into a distance matrix and will perform bottom-up grouping to generate the clusters (define the relationship between data points), it is wise to carefully choose which distance calculation and linkage method are optimal for evaluating the data.  Beside considering what type of question will be asked (does co-expression or expression magnitude provide a better measure for analysis?), a valuable measure to evaluate how well the hierarchical structure represents the data is the cophenetic correlation coefficient.  However, as a single metric masks the nuances of the structure in relationship the the data - i.e. a similar correlation values between two different parameter settings does not reflect a similar structure.  To mitigate this short coming, the correlation between these structures was evaluated.  Linkage methods can be rationally chosen but require validation by looking at the dendrogram.  These tools help reduce the amount of structures needed to be investigated and provide a quantitative means of assessment.  Reviewing the dendrogram after the most appropriate parameters were chosen there seems to be three major groups with 2-4 isolated groups.  To evaluate the height at which the tree needs to be cut to capture most of the biology, an iterative function evaluates how well the leaves separate into the different groups defined by the biology of interest over a range of leaf counts.  Reviewing the results, the grouping of the samples to biological feature begins at leaf count 6 and does not improve with increased leaf counts.  This leaf count is validated with another visualization of the dendrogram colored by leaf.  Despite the success of accurately separating a large majority of the samples, the cluster count (6) was double that of the group count (3).  Hypothetically, if the results were worse the information would not provide sufficient confidence that the structure within the data is relevant to the biology of interest.  Should these isolated groups/samples be considered outliers? Erros in the experiment? This is where dimnensionality reduction comes in handy.  The goal is to help remove some of the unwanted noise in the data, reinforce the presence of any global/local outliers, and hopefully reveal a stronger relationship between the data structure and the biology.  

## Common Dimensionality Reduction Methods

Now, lets look at several different DR tools for the following DR techniques: PCA, tSNE, UMAP and MDS. 
**Important Note: unless a specific seed is set tSNE and UMAP will yield different results per run, which will cause some degree of discrepancy between optimal hyper parameter settings

### EA2 Code Chunk

``` {r EA2, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, class.source= "fold-show", fig.height=5, fig.width=8, results='show', cache=FALSE}
# reset graphical parameters 
par(old_par)

# PCA
pca_vst = optimizePCA(data_object = RNAseqCounts_vst, feature_labels = list(c(sample_metaData$infection),c(sample_metaData$ancestry)), total_variance = 0.8)
pca_vst = optimizePCA(data_object = RNAseqCounts_vst, feature_labels = c(sample_metaData$infection_ancestry), total_variance = 0.8)
plotPCA(pca_object = pca_vst, feature_labels = c(sample_metaData$infection_ancestry), axes = c(1,4), axes_labels = c("PC1", "PC4"))

par(old_par)
# tSNE hyper parameter tuning
optimizeTSNE(data_object = t(RNAseqCounts_vst), feature_labels = c(sample_metaData$infection_ancestry), usage = "perplexity")
optimizeTSNE(data_object = t(RNAseqCounts_vst), feature_labels = c(sample_metaData$infection_ancestry), usage = "iterations", neighbors = 75)
optimizeTSNE(data_object = t(RNAseqCounts_vst), feature_labels = c(sample_metaData$infection_ancestry), usage = "check", neighbors = 75, iterations = 6*50)
tSNE_samples = Rtsne(t(RNAseqCounts_vst), dims = 2, perplexity = 75, max_iter = 6*50, pca = TRUE, theta = 0.25, check_duplicates = FALSE)

par(old_par)
# UMAP hyper parameter "tuning"
set.seed(1)
optimizeUMAP(data_object = t(RNAseqCounts_vst), feature_variable = c(sample_metaData$infection_ancestry), usage = "neighbors")
optimizeUMAP(data_object = t(RNAseqCounts_vst), feature_variable = c(sample_metaData$infection_ancestry), usage = "distance", neighbors = c(50))
optimizeUMAP(data_object = t(RNAseqCounts_vst), feature_variable = c(sample_metaData$infection_ancestry), usage = "iterations", neighbors = c(50), min_distance = 0.25)
optimizeUMAP(data_object = t(RNAseqCounts_vst), feature_variable = c(sample_metaData$infection_ancestry), usage = "check", neighbors = c(50), min_distance = 0.25, iterations = 300, display_legend = TRUE)
umap_samples = umap(t(RNAseqCounts_vst), method = "naive", n_neighbors = 50, n_components = 2, min_dist = 0.25, n_epoch = 300, metric = "euclidean")

par(old_par)
optimizeUMAP(data_object = pca_vst$x[,1:6], feature_variable = c(sample_metaData$infection_ancestry), usage = "neighbors")
optimizeUMAP(data_object = pca_vst$x[,1:6], feature_variable = c(sample_metaData$infection_ancestry), usage = "distance", neighbors = c(50))
optimizeUMAP(data_object = pca_vst$x[,1:6], feature_variable = c(sample_metaData$infection_ancestry), usage = "iterations", neighbors = c(50), min_distance = 0.25)
optimizeUMAP(data_object = pca_vst$x[,1:6], feature_variable = c(sample_metaData$infection_ancestry), usage = "check", neighbors = c(50), min_distance = 0.3, iterations = 100, display_legend = TRUE)
umap_samples2 = umap(pca_vst$x[,1:6], method = "naive", n_neighbors = 35, n_components = 2, min_dist = 0.3, n_epoch = 200, metric = "euclidean")

par(old_par)
# MDS 
mdsOptimize(data_object = RNAseqCounts_vst, feature_labels = c(sample_metaData$infection_ancestry))
mds_samples = mdsOptimize(data_object = RNAseqCounts_vst, feature_labels = c(sample_metaData$infection_ancestry), data_return = "classic")

par(old_par)
mdsOptimize(data_object = t(pca_vst$x[,1:6]), feature_labels = c(sample_metaData$infection_ancestry))
mds_samples2 = mdsOptimize(data_object = t(pca_vst$x[,1:11]), feature_labels = c(sample_metaData$infection_ancestry), data_return = "classic")

# clear system memory
set.seed(NULL)
rm()
```

### EA2 Discussion

Principal Component Analysis (PCA):
PCA allows for DR through means of decreasing axes of variance within the data structure.  For this DR technique it is important to remove as much "noisy variance" from the data structure and evaluate the distribution of variance among the relevant axes. To capture 80% of the variance, which assumes that 20% of the variance is simply noise within the data, the top 11 PCs approximate 80% of the total variance.  Analysis of the distribution of variance across these 11 PCs reveals that the first 4 PCs contribute significantly more to variance than the remaining 7, which show limited changes to variance contribution between them.  Visualization of the top 3 PCs in 2D plots identifies that the biological feature of interest correlates to the PC 1 and PC3 with the best separation visualized by PC3 vs PC1.  PC2 although contributing significantly more to variance (seen in the screeplot and the 2D visualizations) does not provide separation by the biological feature of interest  This suggests that the variance in this axis is representative of some other feature captured by the structure of the expression data absent of the current biological feature of interest.  In cases like this it would be valuable to investigate other data features in the meta data, especially ones that show moderate to poor correlation with the current biological feature of interest.  It is also interesting to see that the variance contribution difference between PC3 and PC4 is minimal.  To be thorough, a quick 2D PC score plot was generated to evaluate the dimension.  Although the dimension did not generate good separation between the different groups defined by the biological feature of interest, it did however reasonably separate the data into two groups (opposing directions of variance).  Again, suggestive of some potential biological feature but not the one of interest.  With the "optimal" number of PCs to include (11), the PCA scores can be subset and used as input for other DR techniques.

t-Distribution Stochastic Neighbor Embedding (tSNE):
tSNE through Rtsne() has a built in algorithm to perform PCA first.  Although this functionality can be turned off and could be used to demonstrate its value here, it was opted out of since tSNE without PCA is practically never recommended.  It is computationally expensive and tSNE is generally improved by PCA.  Hyper parameter tuning for tSNE focuses on the number of nearest neighbors used for embedding and the maximum iterations needed to reduce dissimilarity.  Besides visualizing the "goodness of fit", the KL divergence is used to evaluate the dissimilarity between the original data and the embedding.  How the data shifts from favoring local to global structure can be observed and finding the balance between the two while visually capturing the ideal groupings is simply a good rationale. As an added bonus, minimizing the max iterations will reduce execution time.   The stabilization we seen across different numbers of iterations is with the groups rather than the data within the reduced dimensional space as a whole.     

Uniform Manifold Approximation and Projection (UMAP):
UMAP through umap() does not perform PCA first unlike Rtsne(), thus it is inputted manually in the second iteration.  Similar to tSNE, hyper parameter tuning is dependent primarily on visualization to capture the data feature groupings as well as try to balance local and global structure influence on the embedding.  Here, hyper parameter tuning is done with number of nearest neighbors, minimum distance between neighbors and maximum number of iterations to stabilize the embedding.  Ultimately, the constancy between maximum number of iteration values communicates a reasonable hyper parameter tuning.   The sequential tuning process quickly stabilizes the embedding of both the data set and PCA scores.  


Multidimensional Scaling (MDS):



PCA was performed to reduce the dimension of the count data set.  Covariance captured within PC1 aligns perfectly with the experimental treatment.  PC2 captures variance in a dimension that does not exhibit any alignment with infection type while PC3 does well at separating infection from non-infection samples, suggesting that the variance in this dimension has some sort of biological relevance in line with the experiment.  A screeplot identifies that the overwhelming majority of the covariance within the data is captured within the first 4 PCs.  After truncating the data along with the dimensional reduction of PCA (use of PCs 1-4), outliers from previous hierarchical clustering by samples were avoided and were able to cluster the groups to experimental infection treatments with 100% accuracy and reflect the PCA results interpretation. Other dimensional reduction techniques are viable for reducing the noise and improving the expression bias.  To follow up, t-SNE and UMAp methods will be optimized and then evaluated.  

# Unsupervised Learning (Exploratory Analysis): Clustering 

The importance of clustering in unsupervised learning is simple, an algorithm is applied to group the data by their structural similarities absent of any predefined labels.  By the relationships between each data point and its neighbors, the algorithm captures structure within the data that may or may not corroborate the biological feature/s of interest.  When cluster results identically or closely match groupings by feature, confidence is instilled that the hypothesized relationship has validity. It also can provide insight into how data similarly contributes to the data structure when no such defining feature exists for them, either achieved by looking at individual species rather than features or when the current feature/s of interest show very weak correlation with the clustering results.  At this point it should be clear that clustering is a powerful driver for hypothesis validation and generation.  

Now, the following clustering methods will be demonstrated: the previous discussed method (Hierarchical), K-means (centroid-based) and DBSCAN (density-based) clustering. Like DR techniques, unique parameter tuning is essential to accurately apply the clustering algorithm to the data.

## Hierarchical Clustering

### EA3 Code Chunk

``` {r EA3, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, class.source= "fold-show", fig.height=5, fig.width=8, results='show', cache=FALSE}

par(old_par) # reset graphic parameters

# PCA Hierarchical Clustering
cophTable_pca = analysisCophenetic(data_object = pca_vst$x[,1:6], feature_labels = c(sample_metaData$infection_ancestry), usage = "reduced")
cophTable_pca
analysisHCLUSTCorrelation(data_object = pca_vst$x[,1:6], feature_labels = c(sample_metaData$infection_ancestry), usage = "reduced", dist_calculation = c("euclidean", "manhattan"), linkage_methods = c("ward.D2","single","complete","average","mcquitty","centroid"))
analysisHCLUSTCorrelation(data_object = pca_vst$x[,1:6], feature_labels = c(sample_metaData$infection_ancestry), usage = "reduced", dist_calculation = c("euclidean", "pearson"), linkage_methods = c("ward.D2","single","complete","average","mcquitty","centroid"))

### Average linkage method selected for magnitutde and coepxression evaluation
dend_pca_magnitude = objectHCLUST(data_object = pca_vst$x[,1:6], feature_labels = c(sample_metaData$infection_ancestry), usage = "reduced", return_object = "dend", dend_plot = FALSE) # view dendrogram of optimal distance calculation and linkage method and create dendrogram object
optimizeTree(hc_object = dend_pca_magnitude, leaf_count = 10, feature_labels = c(sample_metaData$infection_ancestry)) # evaluate cutting of the structure into a set nu8mber of groups and see how well it clusters the data by feature
plot(color_branches(dend_pca_magnitude, k = 4)) # view dendrogram with the optimal number of leaves to capture the biology

dend_pca_coexpression = objectHCLUST(data_object = pca_vst$x[,1:6], feature_labels = c(sample_metaData$infection_ancestry), usage = "reduced", dist_calculation = "pearson", return_object = "dend", dend_plot = FALSE) # view dendrogram of optimal distance calculation and linkage method and create dendrogram object
optimizeTree(hc_object = dend_pca_coexpression, leaf_count = 10, feature_labels = c(sample_metaData$infection_ancestry)) # evaluate cutting of the structure into a set nu8mber of groups and see how well it clusters the data by feature
plot(color_branches(dend_pca_coexpression, k = 3)) # view dendrogram with the optimal number of leaves to capture the biology

## tSNE Hierarchical Clustering
cophTable_tsne = analysisCophenetic(data_object = tSNE_samples$Y, feature_labels = c(sample_metaData$infection_ancestry), usage = "reduced")
cophTable_tsne
analysisHCLUSTCorrelation(data_object = tSNE_samples$Y, feature_labels = c(sample_metaData$infection_ancestry), usage = "reduced", dist_calculation = c("euclidean", "manhattan"))
analysisHCLUSTCorrelation(data_object = tSNE_samples$Y, feature_labels = c(sample_metaData$infection_ancestry), usage = "reduced", dist_calculation = c("pearson","spearman"))
analysisHCLUSTCorrelation(data_object = tSNE_samples$Y, feature_labels = c(sample_metaData$infection_ancestry), usage = "reduced", dist_calculation = c("euclidean", "pearson"), linkage_methods = c("ward.D","ward.D2","average","centroid"))

### Average linkage method selected for magnitutde and coepxression evaluation
dend_tsne_magnitude = objectHCLUST(data_object = tSNE_samples$Y, feature_labels = c(sample_metaData$infection_ancestry), usage = "reduced", dist_calculation = "euclidean", linkage_method = "average", return_object = "dend", dend_plot = FALSE) # view dendrogram of optimal distance calculation and linkage method and create dendrogram object
optimizeTree(hc_object = dend_tsne_magnitude, leaf_count = 6, feature_labels = c(sample_metaData$infection_ancestry)) # evaluate cutting of the structure into a set nu8mber of groups and see how well it clusters the data by feature
plot(color_branches(dend_tsne_magnitude, k = 3)) # view dendrogram with the optimal number of leaves to capture the biology

dend_tsne_coexpression = objectHCLUST(data_object = tSNE_samples$Y, feature_labels = c(sample_metaData$infection_ancestry), usage = "reduced", dist_calculation = "pearson", linkage_method = "average", return_object = "dend", dend_plot = FALSE) # view dendrogram of optimal distance calculation and linkage method and create dendrogram object
optimizeTree(hc_object = dend_tsne_coexpression, leaf_count = 10, feature_labels = c(sample_metaData$infection_ancestry)) # evaluate cutting of the structure into a set nu8mber of groups and see how well it clusters the data by feature
plot(color_branches(dend_tsne_coexpression, k = 2)) # view dendrogram with the optimal number of leaves to capture the biology

## UMAP Hierarchical Clustering

cophTable_umap = analysisCophenetic(data_object = umap_samples2$layout, feature_labels = c(sample_metaData$infection_ancestry), usage = "reduced")
cophTable_umap
analysisHCLUSTCorrelation(data_object = umap_samples2$layout, feature_labels = c(sample_metaData$infection_ancestry), usage = "reduced", dist_calculation = c("euclidean", "manhattan"))
analysisHCLUSTCorrelation(data_object = umap_samples2$layout, feature_labels = c(sample_metaData$infection_ancestry), usage = "reduced", dist_calculation = c("pearson","spearman"))
analysisHCLUSTCorrelation(data_object = umap_samples2$layout, feature_labels = c(sample_metaData$infection_ancestry), usage = "reduced", dist_calculation = c("euclidean", "pearson"), linkage_methods = c("complete","ward.D2","average","mcquitty","centroid"))

### Average linkage method selected for magnitutde and coepxression evaluation
dend_umap_magnitude = objectHCLUST(data_object = umap_samples2$layout, feature_labels = c(sample_metaData$infection_ancestry), usage = "reduced", dist_calculation = "euclidean", linkage_method = "average", return_object = "dend", dend_plot = FALSE) # view dendrogram of optimal distance calculation and linkage method and create dendrogram object
optimizeTree(hc_object = dend_umap_magnitude, leaf_count = 6, feature_labels = c(sample_metaData$infection_ancestry)) # evaluate cutting of the structure into a set nu8mber of groups and see how well it clusters the data by feature
plot(color_branches(dend_umap_magnitude, k = 3)) # view dendrogram with the optimal number of leaves to capture the biology

dend_umap_coexpression = objectHCLUST(data_object = umap_samples2$layout, feature_labels = c(sample_metaData$infection_ancestry), usage = "reduced", dist_calculation = "pearson", linkage_method = "average", return_object = "dend", dend_plot = FALSE) # view dendrogram of optimal distance calculation and linkage method and create dendrogram object
optimizeTree(hc_object = dend_umap_coexpression, leaf_count = 6, feature_labels = c(sample_metaData$infection_ancestry)) # evaluate cutting of the structure into a set nu8mber of groups and see how well it clusters the data by feature
plot(color_branches(dend_umap_coexpression, k = 2)) # view dendrogram with the optimal number of leaves to capture the biology

## MDS Hierarchical Clustering

cophTable_mds = analysisCophenetic(data_object = mds_samples2$points, feature_labels = c(sample_metaData$infection_ancestry), usage = "reduced")
cophTable_mds
analysisHCLUSTCorrelation(data_object = mds_samples2$points, feature_labels = c(sample_metaData$infection_ancestry), usage = "reduced", dist_calculation = c("pearson","spearman"))
analysisHCLUSTCorrelation(data_object = mds_samples2$points, feature_labels = c(sample_metaData$infection_ancestry), usage = "reduced", dist_calculation = c("euclidean", "manhattan"))
analysisHCLUSTCorrelation(data_object = mds_samples2$points, feature_labels = c(sample_metaData$infection_ancestry), usage = "reduced", dist_calculation = c("euclidean", "pearson"), linkage_methods = c("centroid","complete","average","ward.D2","mcquitty"))

### Average linkage method selected for magnitutde and coepxression evaluation
dend_mds_magnitude = objectHCLUST(data_object = umap_samples2$layout, feature_labels = c(sample_metaData$infection_ancestry), usage = "reduced", dist_calculation = "euclidean", linkage_method = "average", return_object = "dend", dend_plot = FALSE) # view dendrogram of optimal distance calculation and linkage method and create dendrogram object
optimizeTree(hc_object = dend_mds_magnitude, leaf_count = 6, feature_labels = c(sample_metaData$infection_ancestry)) # evaluate cutting of the structure into a set nu8mber of groups and see how well it clusters the data by feature
plot(color_branches(dend_mds_magnitude, k = 3)) # view dendrogram with the optimal number of leaves to capture the biology

dend_mds_coexpression = objectHCLUST(data_object = umap_samples2$layout, feature_labels = c(sample_metaData$infection_ancestry), usage = "reduced", dist_calculation = "pearson", linkage_method = "average", return_object = "dend", dend_plot = FALSE) # view dendrogram of optimal distance calculation and linkage method and create dendrogram object
optimizeTree(hc_object = dend_mds_coexpression, leaf_count = 6, feature_labels = c(sample_metaData$infection_ancestry)) # evaluate cutting of the structure into a set nu8mber of groups and see how well it clusters the data by feature
plot(color_branches(dend_mds_coexpression, k = 2)) # view dendrogram with the optimal number of leaves to capture the biology


```

### EA3 Discussion

pca: euclidean average is highly correlated with all other combinations with high cophenetic values

tSNE: all pearson and spearman combinations are perfectly correlated (positive) and all euclidean and manhattan combinations are positively corealted perfectly and between them is exactly 0.5 positive correlation

UMAP: similar results to tSNE except single higher correlation between the distance and correlation as distance combinations 

MDS: needs review

The most important aspect of this demonstration is the affect of DR on the hierarchical construction of the data.  Regardless of framing the analysis against expression magnitude or expression pattern (coexpression), hierarchical clustering using PCA scores generates the same global structure between the data features with differences confined to the less obvious structures at the lowest levels (height in the dendrogram).  With sample feature sizes this large, the structure at this level is typically overlooked and therefore relationships between individual features within a cluster require a deeper or horizontal analysis method.  Ultimately, it is clear PCA has improved hierarchical clustering of the data to near perfect accuracy.  Using tSNE, UMAP or MDS has a markedly different effect on the analytical pipeline. Ultimately, these DR techniques strip the data of a structure that is effectively analyzed for coexpression. PCA retains this capability, because it is a linear technique capturing the variance across the number of dimensional axes. Once the other three DR techniques embed the data into the new dimensional space, physical distance between them becomes the only true calculation method to construct hierarchical order of similarity.  It may be best to think of it as expression magnitude and pattern have been embedded together within the new structure.  To this end, applying a distance calculation to any of these DR data structures provides the cleanest and perfectly accurate clustering against the biology of interest.  By hierarchical clustering and with the aid of either PCA, tSNE, UMAP or MDS DR, gene expression of macrophages being well characterized by infection response is strongly supported.  Any related hypotheses stemming from this identified relationship would be reasonable to investigate considering this evidence - e.g. the innate immune response of macrophages is unique to the insulting pathogen.  

## Centroid-based Clustering: K-means

### EA4 Code Chunk

``` {r EA4, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, class.source= "fold-show", fig.height=5, fig.width=8, results='show', cache=FALSE}

par(old_par) # reset graphic parameters

# PCA K-means Clustering
optimizeKmeans(data_object = pca_vst$x[,1:6])
# tSNE K-mean Clustering
optimizeKmeans(data_object = tSNE_samples$Y)
# UMAP K-means Clustering
optimizeKmeans(data_object = umap_samples2$layout)
# MDS K-means Clustering
optimizeKmeans(data_object = mds_samples2$points)

```

### EA4 Discussion

This optimization tool for K-means clustering does several things: provide a visualization of the distribution of variance within each cluster (withinSS) by center count and the total change in within (withinSS), between(betweenSS) and ratioed (betweenSS/totalSS) variance by center count.  Ideally, the number of centers is optimized for a data set when the variance within each cluster is consistent and the change in the different variance metrics is minimal with subsequent increases in centers (elbow plots).  

In this case with respect to the biology of interest, optimization of k-means clustering for the data after PCA and MDS do not show promise in providing corroborating evidence.  The the most consistent distribution of withinSS across clusters is less obvious and the elbow plots potentially suggest at least double the number of feature groups is the optimal number of clusters.  Compare these results to the tSNE and umap DR data.  These cases display a sharp contrast.  The results from optimization with these data objects are a clear example of ideal: the withinSS is obvious most consistent with 3 centers and the the elbow plots support 3 centers as the kink before the change in variance metrics is negligible with increases in clsuter count. 

## Density-based Clustering: DBSCAN

### EA5 Code Chunk

``` {r EA5, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, class.source= "fold-show", fig.height=5, fig.width=8, results='show', cache=FALSE}

# PCA DBSCAN 
optimizeDBSCAN(data_object = pca_vst, feature_labels = c(sample_metaData$infection_ancestry), data_type = "pca", usage = "tune")
optimizeDBSCAN(data_object = pca_vst, feature_labels = c(sample_metaData$infection_ancestry), data_type = "pca", usage = "test", n_size = c(16,20,20,35), hp_check)
optimizeDBSCAN(data_object = pca_vst, feature_labels = c(sample_metaData$infection_ancestry), data_type = "pca", usage = "check", hp_check = c(20,3)) # tuned to focus more on local structure

# tSNE DBSCAN 
optimizeDBSCAN(data_object = tSNE_samples, feature_labels = c(sample_metaData$infection_ancestry), data_type = "tsne", usage = "tune")
optimizeDBSCAN(data_object = tSNE_samples, feature_labels = c(sample_metaData$infection_ancestry), data_type = "tsne", usage = "test", n_size = c(0.4,0.5,0.7,1.25), hp_check, sample_replace = FALSE)
optimizeDBSCAN(data_object = tSNE_samples, feature_labels = c(sample_metaData$infection_ancestry), data_type = "tsne", usage = "check", hp_check = c(2,1.25)) # tuned to more global structure over local 

# UMAP DBSCAN
optimizeDBSCAN(data_object = umap_samples2, feature_labels = c(sample_metaData$infection_ancestry), data_type = "umap", usage = "tune")
optimizeDBSCAN(data_object = umap_samples2, feature_labels = c(sample_metaData$infection_ancestry), data_type = "umap", usage = "test", n_size = c(0.3,0.35,0.4,0.85), hp_check, sample_replace = FALSE)
optimizeDBSCAN(data_object = umap_samples2, feature_labels = c(sample_metaData$infection_ancestry), data_type = "umap", usage = "check", hp_check = c(0.3,3))
optimizeDBSCAN(data_object = umap_samples2, feature_labels = c(sample_metaData$infection_ancestry), data_type = "umap", usage = "check", hp_check = c(0.35,4))
optimizeDBSCAN(data_object = umap_samples2, feature_labels = c(sample_metaData$infection_ancestry), data_type = "umap", usage = "check", hp_check = c(0.85,20))

# MDS DBSCAN
optimizeDBSCAN(data_object = mds_samples2, feature_labels = c(sample_metaData$infection_ancestry), data_type = "mds", usage = "tune")
optimizeDBSCAN(data_object = mds_samples2, feature_labels = c(sample_metaData$infection_ancestry), data_type = "mds", usage = "test", n_size = c(10,11,14,25), hp_check, sample_replace = FALSE)
optimizeDBSCAN(data_object = mds_samples2, feature_labels = c(sample_metaData$infection_ancestry), data_type = "mds", usage = "check", hp_check = c(14,6))

```

### EA5 Discussion

Hyper parameter tuning for DBSCAN's density-based clustering leverages the "dbscan" package, specifically the kNNdistplot() function.  It allows us to visualize the relationship between minimum number of neighbors for recognizing a neighborhood and the size of the neighborhood vs number of points captured.  Interruption of these plots allows use to find the optmial neighborhood size per neighborhood point minimum, which translates to defining clusters by high density relative to these values.  If it isn't obvious already, the spatial relationship between data within its defined dimensional space jointly and ultimately defines these hyper parameter values.  In other words, the Dr technique and therefore how the high dimensional data is embedded into the low dimensional state in junction with the feature of interest strongly influences the tuning process.  Thus, it is crucial to evaluate a density-based clustering algorithm like dbscan across several possible optimal settings and compare.  The lack of actively deciding how many centers/clusters are recognized lets the algorithm speak for itself.  Examples of its execution demonstrate that the effects of favoring either end of the spectrum relative to the low dimensional data (see output for UMAP examples).  Additionally, the lack of need to define the number of clusters/centers allows for the algorithm to assign data points to a "0" cluster, which is interrupted as an anomaly within the data.  It certainly saves time performing global and local anomaly detection methods, which will be covered later.  

# Unsupervised Learning: Application for Species Identification

So far unsupervised learning hasn't extended beyond a global level analysis of the data.  Its application has been limited to anomaly detection and validation of the biological question/s at hand.  This, however, is not the scope of its application.  It is also excellent at discovering key species within the data underlining the biology of interest.  Considering its main attractive feature, species identified with these methods offer a unique confidence in their significance yet still require skepticism nonetheless.   

## Principal Component Analysis At the Core

Because PCA is a linear technique, the contribution of species to the variance at the feature level can be traced.  This aspect alone makes PCA an extremely powerful tool and the backbone of most unsupervised learning pipelines for gene expression analysis.  Not only does the output therefore provide insights for interpretation at the species level but also can serve as input for the other DR techniques with subsequent to clustering, same as the feature analysis.  The connection between the DR of the features and species is maintained, thus provides a beautifully complimentary analytic approach. Lets first look at two PCA analysis tools before following up with a complete unsupervised learning pipeline to discover species of interest within the data.  A comparison of the results may reveal why or why not PCA is enough.  

### EA6 Code Chunk

``` {r EA6, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, class.source= "fold-show", fig.height=5, fig.width=8, results='show', cache=FALSE}
par(old_par)
IS_PCA = analysisPCA(data_object = RNAseqCounts_vst, feature_labels = c(sample_metaData$infection_ancestry), plot_data = TRUE, data_return = TRUE, load_threshold = 0.05)
remove_samples = c(492,491,494,484,489)
IS_PCA_QC = analysisPCA(data_object = RNAseqCounts_vst[,-remove_samples], feature_labels = c(sample_metaData$infection_ancestry)[-remove_samples], plot_data = TRUE, data_return = TRUE, load_threshold = 0.05)
# identified species ("IS") set to a loading threshold of 0.05 across the top 3 PCs
IS_PCA 
IS_PCA_QC

# identify important species using the distribution of contribution to variance across species
target_sepcies_base = filterLoadings(pca_object = pca_vst, species_labels = rownames(RNAseqCounts_vst), npcs = 6)
length(target_sepcies_base)
target_species_directional = filterLoadings(pca_object = pca_vst, species_labels = rownames(RNAseqCounts_vst), npcs = 6, nd_return = FALSE, top_return = FALSE)
length(target_species_directional)
target_species_nondirectional = filterLoadings(pca_object = pca_vst, species_labels = rownames(RNAseqCounts_vst), npcs = 6, d_return = FALSE, top_return = FALSE)
length(target_species_nondirectional)
target_species_top = filterLoadings(pca_object = pca_vst, species_labels = rownames(RNAseqCounts_vst), npcs = 6, d_return = FALSE, nd_return = FALSE, top_pc = c(1,3))
length(target_species_top)
target_species_counter = filterLoadings(pca_object = pca_vst, species_labels = rownames(RNAseqCounts_vst), npcs = 6, d_return = FALSE, nd_return = FALSE, top_pc = c(2))
length(target_species_counter)

target_species_top
target_species_counter

# clear Memory
rm()
```

### EA6 Discussion

analysisPCA: 
This comprehensive tool offer several different features, removal the the "noise" variance (previous found), detection of global and local anomalies using three different metrics, analysis of species contributions to identified "true" variance, and final comparative visualizations to aid in the analysis.  Multiple metrics for anomaly detection are included to account for global and local detection as well as these methods are not universally absolute.  Reconstruction errors and local outlier factors are easy to interpret only when extreme.  Mahalanobis distance is useful but only when the structure of the data within the dimensions of interest fits a general circular shape and loses its usefulness when data clusters in irregular or elliptical shapes.  Some of these of issues can be mitigated by visualizing anomaly placement within the dimensional space but this isn't always the case.  That being said we are able to recognize how samples identified with higher LOF may indeed be anomalies worth removing based on the visualization of the first 3 PCs.  A rationale could be to remove any samples that reasonably breaks from the observable pattern of the feature groups from these plots.  This would be a very conservative approach to sample QC and acceptable considering how many samples there are per group and how few the identified anomalies are.  Run a second time with identified samples removed (index: ) we can see the effect.  Because of how these anomaly detection metrics are calculated and identified, subsequent execution will identify new samples as potential outliers.  Although an iterative process does reveal information about the structure of the data, it is only recommended for advanced users with specific goals in mind.  

The species identification using this tool is also very conservative to its contribution to the structure of the data.  It is limited to just the first 3 (top) PCs of the data and a specific absolute load threshold value is chosen.  This threshold is chosen by the user and therefore most strongly influenced by the structure of the PC loading plots rather than a statistical measure.  The dependency on the visualization itself as well as assumptions made by the user makes it both powerful and potentially misleading.  Although in practice this should result in limited identification of species and they will be major contributors to the top 3 principal components, the dimensions housing the greatest variance of the data, the problem is this approach to it is not adhering to an "unsupervised" approach.  It is more accurately a supervised last step to determining the species of interest, which is not inherently wrong but break away from the overall approach itself.  To approach species identification more true unsupervised learning the next tool is better suited, "filterLoadings()". 

filterLoadings:
This tool's functionality includes capturing the species with the most relevant directional or non-directional average contributions to variance as well as the targeting a subset of principal components of interest for a total of three different criteria applied to extract species of interest. Each are controlled independently along withe their distribution thresholds.  Although the same supervision could be applied here, these threshold can be set to capture species fitted to a specified probability rather than the look of the distribution. Because different schools of thought place different priority on directional vs non-directional loading as well as the value of targeting specific PCA dimensions, the level of control over the the direction of the unsupervised process is powerful for analysis.  Several different avenues can be explored downstream and validated.  A demonstration of this usefulness the function's functionality is with the two returned species lists from PCs 1 and 3 vs PC 2.  A quick search from these results shows the species identified from PC2 are not obviously related to the biology of interest compared to the species returned from PC 1 and 3, which matches the PC scores presented by the analysisPCA function. 

## Leveraging the Lineararity of PCA: An Example using UMAP

### EA7 Code Chunk

``` {r EA7, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, class.source= "fold-show", fig.height=5, fig.width=8, results='show', cache=FALSE}

par(old_par)
optimizeUMAP(data_object = pca_vst$rotation[,1:6], feature_variable = c(rownames(RNAseqCounts_vst)), usage = "neighbors")
optimizeUMAP(data_object = pca_vst$rotation[,1:6], feature_variable = c(rownames(RNAseqCounts_vst)), usage = "distance", neighbors = c(20))
optimizeUMAP(data_object = pca_vst$rotation[,1:6], feature_variable = c(rownames(RNAseqCounts_vst)), usage = "iterations", neighbors = c(20), min_distance = 0.1)
optimizeUMAP(data_object = pca_vst$rotation[,1:6], feature_variable = c(rownames(RNAseqCounts_vst)), usage = "check", neighbors = c(15), min_distance = 0.1, iterations = 1000, display_legend = FALSE)
optimizeUMAP(data_object = pca_vst$rotation[,1:6], feature_variable = c(rownames(RNAseqCounts_vst)), usage = "check", neighbors = c(10), min_distance = 0.1, iterations = 1000, display_legend = FALSE)

optimizeUMAP(data_object = pca_vst$rotation[,1:6], feature_variable = c(rownames(RNAseqCounts_vst)), usage = "distance", neighbors = c(5))
optimizeUMAP(data_object = pca_vst$rotation[,1:6], feature_variable = c(rownames(RNAseqCounts_vst)), usage = "iterations", neighbors = c(5), min_distance = 0.25)
optimizeUMAP(data_object = pca_vst$rotation[,1:6], feature_variable = c(rownames(RNAseqCounts_vst)), usage = "check", neighbors = c(5), min_distance = 0.05, iterations = 500, display_legend = FALSE)            
umap_species = umap(pca_vst$rotation[,1:6], method = "naive", n_neighbors = 5, n_components = 2, min_dist = 0.05, n_epoch = 500, metric = "euclidean")

```

### EA7 Discussion

First pass hyper parameter tuning had unexpected results.  Increasing the iterations improves embedding at the cost of speed, which typically results in tighter clusters. When this doesn't occur it suggest an issue with the other hyper parameter settings.  Second pass shows high stability across the range of minimum distances and iterations but not perfect.  Higher iterations resulted in a loss of separation a grouping a species but a single species remains separated.  Although separation was lost, a density-based clustering method may still catpure the separation between them.  

### EA8 Code Chunk

``` {r EA8, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, class.source= "fold-show", fig.height=5, fig.width=8, results='show', cache=FALSE}

optimizeDBSCAN(data_object = umap_species, feature_labels = c(rownames(RNAseqCounts_vst)), data_type = "umap", usage = "tune")
optimizeDBSCAN(data_object = umap_species, feature_labels = c(rownames(RNAseqCounts_vst)), data_type = "umap", usage = "test", n_size = c(0.1,0.125,0.2,0.2), hp_check, sample_replace = FALSE, plot_legend = FALSE)
optimizeDBSCAN(data_object = umap_species, feature_labels = c(rownames(RNAseqCounts_vst)), data_type = "umap", usage = "check", hp_check = c(0.2,20))

par(old_par)
optimizeKmeans(data_object = umap_species$layout)

par(old_par)
cophTable_species = analysisCophenetic(data_object = umap_species$layout, feature_labels = c(rownames(RNAseqCounts_vst)), usage = "reduced")
cophTable_species

dend_umap_species = objectHCLUST(data_object = umap_species$layout, feature_labels = c(rownames(RNAseqCounts_vst)), usage = "reduced", dist_calculation = "euclidean", linkage_method = "average", return_object = "dend", dend_plot = FALSE)
optimizeTree(hc_object = dend_umap_species, leaf_count = 10, feature_labels = c(rownames(RNAseqCounts_vst))) # evaluate cutting of the structure into a set nu8mber of groups and see how well it clusters the data by feature
species_hclustering = cutree(dend_umap_species, k = 5)
plot(color_branches(dend_umap_species, k = 5)) # view dendrogram with the optimal number of leaves to capture the biology
table(x)

```


### EA8 Discussion

Because of the spatial embedding of the data within the reduced dimensional space, K-means clustering is unlikely the best method.  Judging by the amount of overlay between the species within the space, density-based clustering may best capture subtleties unseen by the previous visualization.   

## Generating Hypotheses from Exploratory Analysis Applied to the GEO Data Set

### EA9 Code Chunk

``` {r EA9, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, class.source= "fold-show", fig.height=5, fig.width=8, results='show', cache=FALSE}

species_DBclustering = dbscan(umap_species$layout, eps = 0.2, minPts = 20)

table(species_DBclustering$cluster)
table(species_hclustering)
DataSubset_List = list() # create a list of Data frames for subsets of RNAseqCounts_vst for supervised learning
subset_index = 1
# flow control to make data subsets from each clustering results
for (cluster in unique(species_DBclustering$cluster)) {
  index_target = which(species_DBclustering$cluster == cluster)
  subset_data = as.data.frame(RNAseqCounts_vst[index_target,])
  DataSubset_List[[subset_index]] = subset_data
  subset_index = subset_index + 1
}
for (cluster in unique(species_hclustering)) {
  index_target = which(species_hclustering == cluster)
  subset_data = as.data.frame(RNAseqCounts_vst[index_target,])
  DataSubset_List[[subset_index]] = subset_data
  subset_index = subset_index + 1
}

IS_names_top = rownames(IS_PCA[abs(IS_PCA$PC1) >= 0.05 | abs(IS_PCA$PC2) >= 0.05 | abs(IS_PCA$PC3) >= 0.05,])
ISqc_names_top = rownames(IS_PCA_QC[abs(IS_PCA_QC$PC1) >= 0.05 | abs(IS_PCA_QC$PC2) >= 0.05 | abs(IS_PCA_QC$PC3) >= 0.05,])
table(IS_names_top == ISqc_names_top)
length(IS_names_top)

DataSubset_List[[subset_index]] = as.data.frame(RNAseqCounts_vst[IS_names_top,])
subset_index = subset_index + 1

```


### EA9 Discussion

Discuss the details leading into the list of subset data sets for supervised learning method application. 

Touch on hypothesis generation from exploratory analysis alone and validation with supervised learning (Over Enrishment Analysis, Gene Set Enrichment analysis, network analysis, etc)


# Supervised Learning 

## Differential Expression Analysis

A fundamental supervised learning method is differential expression analysis in which experimental groups are compared gene by gene.  The fold-change in expression between the two defines how the gene's expression changes from one group to the other, both directional and at a relative magnitude. The larger the fold-change magnitude the more the expression of the gene shifts.  P-values aid in the interruption of significance and can be adjusted for multiple hypothesis testing.  By setting a cutoff value for the fold-change and ensuring the probability of encountering such a fold-change reaches an also user defined p-value cutoff, a subset of genes with potential biological significance by the specific group comparison is identified, "differentially expressed genes" (DEG), and poised for downstream hypothesis generation.  A robust R bioconductor R package exists for this process, DESeq2, which contains the VST normalization method used previous. Because the QC and normalization process was performed prior to this approach with custom tools, only the most basic DESeq2 workflow is needed to perform  differential expression analysis for the data set.  

### DEG Code Chunk

``` {r DEG, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, class.source= "fold-show", fig.height=5, fig.width=8, results='show', cache=FALSE}

# Recreate DESeq object to understand the affect Ancestry has on resposne to infection
deSEQ = DESeqDataSetFromMatrix(countData = RNAseqCounts, colData = sample_metaData, design = ~ ancestry + infection + ancestry:infection)# DEseq obejct
dds = DESeq(deSEQ) # run DESeq
plotDispEsts(dds) # View dispersion plot
resultsNames(dds) # get names of coefficients for results data

# What is the effect of ancestry in non-infected samples only?
result_baseline_ancestry = results(dds, name = "ancestry_European_vs_African") 
result_baseline_ancestry_shrink = lfcShrink(dds, res = result_baseline_ancestry, type = "ashr")
plotDESeqResults(data_object = result_baseline_ancestry, "Non-infected African vs European Populations")
plotDESeqResults(data_object = result_baseline_ancestry_shrink, "Shrunk Non-infected African vs European Populations", shrink_data = TRUE, return_data = TRUE)

# Is the response to Listeria infection different in European population?
# difference in Listeria response minus the baseline difference between the ancestries
result_Listeria_ancestry = results(dds, name = "ancestryEuropean.infectionListeria") 
result_Listeria_ancestry_shrink = lfcShrink(dds, res = result_Listeria_ancestry, type = "ashr")
plotDESeqResults(data_object = result_Listeria_ancestry, "Difference Between European and African Population: Listeria")
plotDESeqResults(data_object = result_Listeria_ancestry_shrink, "Difference Between European and African Population (Shrunk): Listeria", shrink_data = TRUE, return_data = TRUE)
# absolute difference in Listeria response (includes the baseline differences)
result_Listeria_ancestry2 = results(dds, contrast=list(c("ancestry_European_vs_African", "ancestryEuropean.infectionListeria")))
result_Listeria_ancestry_shrink2 = lfcShrink(dds, contrast = list(c("ancestry_European_vs_African", "ancestryEuropean.infectionListeria")), type = "ashr")
plotDESeqResults(data_object = result_Listeria_ancestry2, "Absolute Differences Between Eurpoean and African Population: Listeria")
plotDESeqResults(data_object = result_Listeria_ancestry_shrink2, "Absolute Differences Between Eurpoean and African Population: Listeria (Shrunk): Listeria", shrink_data = TRUE, return_data = TRUE)

# Is the response to Salmonella infection different in European population?
# difference in Salmonella response minus the baseline difference between the ancestries
result_Salmonella_ancestry = results(dds, name = "ancestryEuropean.infectionSalmonella") 
result_Salmonella_ancestry_shrink = lfcShrink(dds, res = result_Salmonella_ancestry, type = "ashr")
plotDESeqResults(data_object = result_Salmonella_ancestry, "Difference Between European and African Population: Salmonella")
plotDESeqResults(data_object = result_Salmonella_ancestry_shrink, "Difference Between European and African Population (Shrunk): Salmonella", shrink_data = TRUE, return_data = TRUE)
# absolute difference in Salmonella response (includes the baseline differences)
result_Salmonella_ancestry2 = results(dds, contrast=list(c("ancestry_European_vs_African", "ancestryEuropean.infectionSalmonella")))
result_Salmonella_ancestry_shrink2 = lfcShrink(dds, contrast = list(c("ancestry_European_vs_African", "ancestryEuropean.infectionSalmonella")), type = "ashr")
plotDESeqResults(data_object = result_Listeria_ancestry2, "Absolute Differences Between Eurpoean and African Population: Salmonella")
plotDESeqResults(data_object = result_Listeria_ancestry_shrink2, "Absolute Differences Between Eurpoean and African Population: Listeria (Shrunk): Salmonella", shrink_data = TRUE, return_data = TRUE)

# What genes respond to Listeria infection in the African population?
result_baseline_Listeria = results(dds, name = "infection_Listeria_vs_Non.infected") 
result_baseline_Listeria_shrink = lfcShrink(dds, res = result_baseline_Listeria, type = "ashr")
plotDESeqResults(data_object = result_baseline_Listeria, "African Popualtion Response to Infection: Listeria")
plotDESeqResults(data_object = result_baseline_Listeria_shrink, "African Population Response to Infection (Shrunk): Listeria", shrink_data = TRUE)

# What genes respond to Salmonella infection in the African population?
result_baseline_Salmonella = results(dds, name = "infection_Salmonella_vs_Non.infected")
result_baseline_Salmonella_shrink = lfcShrink(dds, res = result_baseline_Salmonella, type = "ashr")
plotDESeqResults(data_object = result_baseline_Salmonella, "African Popualtion Response to Infection: Salmonella")
plotDESeqResults(data_object = result_baseline_Salmonella_shrink, "African Population Response to Infection (Shrunk): Salmonella", shrink_data = TRUE)

# Investigate the European population
# restrucutre the design
dds$ancestry <- relevel(dds$ancestry, ref = "European")
dds <- DESeq(dds)
resultsNames(dds)

# What genes respond to Listeria infection in the European population?
result_European_Listeria = results(dds, name = "infection_Listeria_vs_Non.infected") 
result_European_Listeria_shrink = lfcShrink(dds, res = result_baseline_Listeria, type = "ashr")
plotDESeqResults(data_object = result_European_Listeria, "African Popualtion Response to Infection: Listeria")
plotDESeqResults(data_object = result_European_Listeria_shrink, "African Population Response to Infection (Shrunk): Listeria", shrink_data = TRUE)

# What genes respond to Salmonella infection in the European population?
result_European_Salmonella = results(dds, name = "infection_Salmonella_vs_Non.infected")
result_European_Salmonella_shrink = lfcShrink(dds, res = result_baseline_Salmonella, type = "ashr")
plotDESeqResults(data_object = result_European_Salmonella, "African Popualtion Response to Infection: Salmonella")
plotDESeqResults(data_object = result_European_Salmonella_shrink, "African Population Response to Infection (Shrunk): Salmonella", shrink_data = TRUE)

```

### DEG Discussion 

Because the biological question aims tom investigate the interaction between ancestry (African vs European) and response to infection the formula must account for both variables (ancestry and infection) as well as an interaction term, thus the formula best suited for the DESeq analysis is "~ ancestry + infection + ancestry:infection:". The African population is set as the reference term to be inline with  genealogical history of humanity.  Non-infected group is also leveled accordingly to be recognized for baseline differentiation calculation.  The most relevant result comparisons are to tackle the baseline differences between European and African populations and their responses to each infection type.  The results if the baseline (non-infected only) differences between the ancestries reveal several gene, "MTRNR2L8", "RETN", "LILRP2", "SERPINB10", "LOC107984551", "NKG7", "USP12-DT", "DPYSL3", "MXRA7", and "IL36RN". Two approaches were employed for investigating response to infection, differentiation absent of their baseline expression differences in non-infected samples and their absolute differences (including the baseline difference).  Both assessments of the differences absent of their baseline differences return no species that pass the standard error correction and Log2FoldChange cutoff, also indicated on the volcano plots.  The p-value histograms also suggest that low probabilities achieved by some of the species may be due more to chance than represent anything statistically significant.  Only four species are found to be differentiated by absolute difference in response to Listeria infection, "RETN", "SERPINB10", "NCCRP1", and "PLAAT3".  These are the same differentiated species returned for the response to Salmonella infection.  For a sanity check, the response to infections within ancestries were also evaluated. It is reasonable to assume that an inflammatory cell like primary macrophages would display an intense and widespread response to the presence of a bacterial pathogen. As expected the number of species and magnitude of their differentiation between non-infected and infection type are large.  


## Linear Regression Models

### GLM1 Code Chunk

``` {r GLM1, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, class.source= "fold-show", fig.height=5, fig.width=8, results='show', cache=FALSE}
```

### GLM1 Discussion 


## Machine Learning Methods

### ML1 Code Chunk

``` {r ML1, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, class.source= "fold-show", fig.height=5, fig.width=8, results='show', cache=FALSE}
```


### ML1 Discussion


# Hypothesis Generation

## Over Enrichment Analysis (OEA)

### OEA1 Code Chunk

``` {r OEA1, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, class.source= "fold-show", fig.height=5, fig.width=8, results='show', cache=FALSE}
```


### OEA1 Discussion


## Gene Set Enrichment Analysis (GSEA)

### GSEA1 Code Chunk

``` {r GSEA1, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, class.source= "fold-show", fig.height=5, fig.width=8, results='show', cache=FALSE}
```


### GSEA1 Discussion















# title

## Code Chunk

``` {r Q3, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, class.source= "fold-show", fig.height=5, fig.width=8, results='show', cache=FALSE}
```

## Discussion

